Forecasting the 2024 US Presidential Election

### 2024-11-03 paper/paper.qmd
Added latest data

# Specify the predictor columns
predictor_columns <- c(
  "What is the highest level of education you have attained? College graduate",
  "What is the highest level of education you have attained? High school or less",
  "Can you please tell me your gender? Female",
  "Can you please tell me your gender? Male",
  "For statistical purposes only, can you please tell me your ethnicity? White or Caucasian",
  "For statistical purposes only, can you please tell me your ethnicity? Hispanic or Latino of any race",
  "Who did you vote for in the 2020 election? Donald Trump",
  "Who did you vote for in the 2020 election? Joe Biden",
  "Do you approve or disapprove of the job Joe Biden is doing as President? Approve",
  "Do you approve or disapprove of the job Joe Biden is doing as President? Disapprove"
)

@@

From this analysis, we identified eight critical swing states: Michigan, Georgia, Nevada, North Carolina, New Hampshire, Wisconsin, Pennsylvania, and Arizona. These states will be central to our election prediction. Although Minnesota emerged as a potential swing state, we excluded it due to the wide range of polling data, as explained in @tbl-final-summary. Additionally, we will consider the national average in our overall prediction.
```{r}
#| label: fig-party-results
#| fig-cap: Example Predictor Variable Visualization
#| echo: false
#| warning: false
#| message: false

@@

# Calculate 95% confidence interval for each unique state in 'DEM' data
dem_ci_summary <- dem_data %>%
  group_by(state) %>%
  summarise(
    mean_pct = mean(pct, na.rm = TRUE),
    sd_pct = sd(pct, na.rm = TRUE),
    n = sum(!is.na(pct)),
    lower_95 = ifelse(n >= 2, mean_pct - qt(0.975, df = n - 1) * (sd_pct / sqrt(n)), NA),
    upper_95 = ifelse(n >= 2, mean_pct + qt(0.975, df = n - 1) * (sd_pct / sqrt(n)), NA)
  ) %>%
  # Filter out intervals where the bounds go below 40 or above 60, and remove rows where mean_pct exceeds 50
  filter(!(lower_95 < 39 | upper_95 > 60) & mean_pct <= 100)

@@

# Calculate 95% confidence interval for each unique state in 'REP' data
rep_ci_summary <- rep_data %>%
  group_by(state) %>%
  summarise(
    mean_pct = mean(pct, na.rm = TRUE),
    sd_pct = sd(pct, na.rm = TRUE),
    n = sum(!is.na(pct)),
    lower_95 = ifelse(n >= 2, mean_pct - qt(0.975, df = n - 1) * (sd_pct / sqrt(n)), NA),
    upper_95 = ifelse(n >= 2, mean_pct + qt(0.975, df = n - 1) * (sd_pct / sqrt(n)), NA)
  ) %>%
  # Filter out intervals where the bounds go below 40 or above 60, and remove rows where mean_pct exceeds 50
  filter(!(lower_95 < 39 | upper_95 > 60) & mean_pct <= 50)

@@

```{r}
#| label: fig-party-extremes
#| fig-cap: Demographic Differences Between Top Red and Blue States
#| echo: false
#| warning: false
#| message: false

library(tidyverse)

# Prepare the poll_summary data frame
poll_summary <- poll_data %>%
  mutate(party_diff = democrat_pct - republican_pct) %>%
  select(-c(2, 3, 4, 6, 7, 8, 9, 11)) %>%  # Remove specific columns by index
  select(-contains('Mode')) %>%             # Remove columns containing 'Mode' in their names
  select(-contains('registration')) %>%     # Remove columns containing 'registration' in their names
  drop_na()                                 # Remove rows with any NA values

# Function to identify top N blue and red states
identify_top_states <- function(data, n) {
  data %>%
    arrange(desc(party_diff)) %>%
    summarise(
      blue_states = list(state[1:n]),  
      red_states = list(state[(nrow(data) - n + 1):nrow(data)])
    )
}

# Function to compute mean differences for all numeric variables
compute_mean_differences <- function(data, blue_states, red_states) {
  numeric_vars <- data %>%
    select(-state, -party_diff) %>%
    select(where(is.numeric)) %>%
    names()

  blue_means <- data %>%
    filter(state %in% blue_states) %>%
    summarise(across(all_of(numeric_vars), mean, na.rm = TRUE))

  red_means <- data %>%
    filter(state %in% red_states) %>%
    summarise(across(all_of(numeric_vars), mean, na.rm = TRUE))

@@

  top_diff_vars %>%
    pivot_longer(
      cols = c("Blue_Mean", "Red_Mean"), 
      names_to = "Group", 
      values_to = "Mean"
    ) %>%
    mutate(Group = recode(Group, "Blue_Mean" = "Republican State", "Red_Mean" = "Democrat State"))
}

# Function to create the comparison plot
create_comparison_plot <- function(plot_data) {
  ggplot(plot_data, aes(
    x = reorder(Variable, -Mean), 
    y = Mean, 
    fill = Group, 
    label = round(Mean, 1)
  )) +
    geom_col(position = position_dodge(width = 0.9)) +
    geom_text(
      position = position_dodge(width = 0.9), 
      vjust = -0.3, 
      size = 3
    ) +
    # Update the colors and labels for the groups
    scale_fill_manual(values = c("Republican State" = "red", "Democrat State" = "blue")) +
    labs(
      title = "Demographic Differences Between Top Red and Blue States",
      x = "Demographic Variable",
      y = "Mean Percentage",
      fill = "State Group"
    ) +
    scale_y_continuous(limits = c(0, 100)) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 4)
    )
}

# Main analysis function
analyze_state_differences <- function(data, top_n_states = 1, top_n_vars = 4) {
  top_states <- identify_top_states(data, n = top_n_states)

  diff_df <- compute_mean_differences(
    data, 
    blue_states = top_states$blue_states[[1]], 
    red_states = top_states$red_states[[1]]
  )

@@

# Run the analysis
analysis_results <- analyze_state_differences(poll_summary, top_n_states = 1, top_n_vars = 5)

@@

column_refs <- list(
  democrat = "democrat_pct",
  republican = "republican_pct",
  highly_educated = "What is the highest level of education you have attained? College graduate",
  lower_educated = "What is the highest level of education you have attained? High school or less",
  female = "Can you please tell me your gender? Female",
  male = "Can you please tell me your gender? Male",
  non_binary_other = "Can you please tell me your gender? Nonbinary or other",
  caucasian = "For statistical purposes only, can you please tell me your ethnicity? White or Caucasian",
  hispanic_latino = "For statistical purposes only, can you please tell me your ethnicity? Hispanic or Latino of any race",
  vote_trump = "Who did you vote for in the 2020 election? Donald Trump",
  vote_biden = "Who did you vote for in the 2020 election? Joe Biden",
  biden_approve = "Do you approve or disapprove of the job Joe Biden is doing as President? Approve",
  biden_disapprove = "Do you approve or disapprove of the job Joe Biden is doing as President? Disapprove",
  previous_vote_trump ="Who did you vote for in the 2020 election? Donald Trump",
  previous_vote_joe ="Who did you vote for in the 2020 election? Joe Biden"

@@

# Calculate poll_summary with rounding to the nearest tenth
poll_summary <- poll_data %>%
  group_by(state) %>%
  summarise(
    dem_total_pct = round(sum(.data[[column_refs$democrat]], na.rm = TRUE), 1),
    rep_total_pct = round(sum(.data[[column_refs$republican]], na.rm = TRUE), 1),
    highly_educated_pct = round(mean(.data[[column_refs$highly_educated]], na.rm = TRUE), 1),
    lower_educated_pct = round(mean(.data[[column_refs$lower_educated]], na.rm = TRUE), 1),
    female_pct = round(mean(.data[[column_refs$female]], na.rm = TRUE), 1),
    male_pct = round(mean(.data[[column_refs$male]], na.rm = TRUE), 1),
    non_binary_other_pct = round(mean(.data[[column_refs$non_binary_other]], na.rm = TRUE), 1),
    caucasian_pct = round(mean(.data[[column_refs$caucasian]], na.rm = TRUE), 1),
    minority_pct = round(100 - caucasian_pct, 1)
  ) %>%
  mutate(party_diff = round(dem_total_pct - rep_total_pct, 1)) %>%
  drop_na()

# Select relevant numeric columns for correlation matrix, excluding PartyDifference
corr_data <- poll_summary %>%
  select(
    Democrat = dem_total_pct,
    Republican = rep_total_pct,
    HighlyEducated = highly_educated_pct,
    LowerEducated = lower_educated_pct,
    Female = female_pct,
    Male = male_pct,
    NonBinary = non_binary_other_pct,
    Caucasian = caucasian_pct,
    Minority = minority_pct
    # No age-related columns included anymore
  )

@@

# Visualize the correlation matrix without hierarchical clustering
ggcorrplot(cor_matrix, 
           hc.order = FALSE,  
           type = "lower", 
           lab = TRUE, 
           lab_size = 2,  # Size for correlation labels
           title = "Correlation Matrix of Predictors and Party Lean",
           colors = c("blue", "white", "red"),  
           legend.title = "Correlation") +
  theme(
    plot.title = element_text(size = 12, hjust = 0.5),  # Title size and centered
    axis.text = element_text(size = 1),  # Axis text size
    legend.text = element_text(size = 10),  # Legend text size
    legend.title = element_text(size = 10),  # Legend title size
    axis.title = element_text(size = 1)  # Axis labels size
  )
```

```{r}
#| label: fig-predictors-party-pct-2
#| fig-cap: Example Voting and Approval Predictor Variable Visualization
#| echo: false
#| warning: false
#| message: false
# Calculate poll_summary with rounding to the nearest tenth
poll_summary_2 <- poll_data %>%
  group_by(state) %>%
  summarise(
    dem_total_pct = round(sum(.data[[column_refs$democrat]], na.rm = TRUE), 1),
    rep_total_pct = round(sum(.data[[column_refs$republican]], na.rm = TRUE), 1),
    vote_trump_pct = round(mean(.data[[column_refs$vote_trump]], na.rm = TRUE), 1),
    vote_biden_pct = round(mean(.data[[column_refs$vote_biden]], na.rm = TRUE), 1),
    biden_approve_pct = round(mean(.data[[column_refs$biden_approve]], na.rm = TRUE), 1),
    biden_disapprove_pct = round(mean(.data[[column_refs$biden_disapprove]], na.rm = TRUE), 1)
  ) %>%
  mutate(party_diff = round(dem_total_pct - rep_total_pct, 1)) %>%
  drop_na()

# Select relevant numeric columns for correlation matrix, excluding PartyDifference
corr_data_2 <- poll_summary_2 %>%
  select(
    Democrat = dem_total_pct,
    Republican = rep_total_pct,
    VotedTrump2020 = vote_trump_pct,
    VotedBiden2020 = vote_biden_pct,
    ApprovesOfBiden = biden_approve_pct,
    DisapprovesOfBiden = biden_disapprove_pct
  )

@@

```{r}
#| label: fig-predictors-education-race-voting
#| fig-cap: Correlation Matrix of Education, Race, and Past Voting Behavior
#| echo: false
#| warning: false
#| message: false

# Calculate poll_summary with rounding to the nearest tenth
poll_summary_3 <- poll_data %>%
  group_by(state) %>%
  summarise(
    dem_total_pct = round(sum(.data[[column_refs$democrat]], na.rm = TRUE), 1),
    rep_total_pct = round(sum(.data[[column_refs$republican]], na.rm = TRUE), 1),
    highly_educated_pct = round(mean(.data[[column_refs$highly_educated]], na.rm = TRUE), 1),
    lower_educated_pct = round(mean(.data[[column_refs$lower_educated]], na.rm = TRUE), 1),
    caucasian_pct = round(mean(.data[[column_refs$caucasian]], na.rm = TRUE), 1),
    minority_pct = round(100 - caucasian_pct, 1),
    vote_trump_pct = round(mean(.data[[column_refs$vote_trump]], na.rm = TRUE), 1),
    vote_biden_pct = round(mean(.data[[column_refs$vote_biden]], na.rm = TRUE), 1)
  ) %>%
  mutate(party_diff = round(dem_total_pct - rep_total_pct, 1)) %>%
  drop_na()

# Select relevant numeric columns for correlation matrix, excluding PartyDifference
corr_data_3 <- poll_summary_3 %>%
  select(
    Democrat = dem_total_pct,
    Republican = rep_total_pct,
    HighlyEducated = highly_educated_pct,
    LowerEducated = lower_educated_pct,
    Caucasian = caucasian_pct,
    Minority = minority_pct,
    VotedTrump2020 = vote_trump_pct,
    VotedBiden2020 = vote_biden_pct
  )

# Calculate the correlation matrix
cor_matrix_3 <- cor(corr_data_3, use = "complete.obs")

# Set correlations less than -0.9 to NA
cor_matrix_3[cor_matrix_3 < -0.85] <- NA

# Visualize the correlation matrix without hierarchical clustering
ggcorrplot(cor_matrix_3, 
           hc.order = FALSE,  
           type = "lower", 
           lab = TRUE, 
           lab_size = 2,  # Size for correlation labels
           title = "Correlation Matrix of Education, Race, and Past Voting Behavior",
           colors = c("blue", "white", "red"),  
           legend.title = "Correlation") +
  theme(
    plot.title = element_text(size = 12, hjust = 0.5),  # Title size and centered
    axis.text = element_text(size = 1),  # Axis text size
    legend.text = element_text(size = 10),  # Legend text size
    legend.title = element_text(size = 10),  # Legend title size
    axis.title = element_text(size = 1)  # Axis labels size
  )

```
The correlation matrix in @fig-predictors-party-pct shows how different predictors relate to party preferences. We can see which factors, like education or gender, are more strongly tied to Democratic or Republican leanings. Blue shading indicates stronger Democratic support, while red shows stronger Republican support. Key observations include higher education being linked to Democratic support and lower education favoring Trump in 2020. Minority groups generally lean Democratic, while Caucasian/White voters tend to support Trump. Past voter behavior shows Biden supporters remain Democratic, and Trump supporters disapprove of Biden. Additionally, women lean slightly more Democratic compared to men, who often lean Republican.

```{r}
#| label: fig-points-over-time
#| tbl-cap: "Parties Approval"
#| echo: false
#| warning: false
#| message: false
# Convert `end_date` to Date format, ensure numeric columns, and limit to 2024 and onward
time_data_graph <- time_data %>%
  mutate(
    end_date = as.Date(end_date, format = "%Y-%m-%d"),  # Format matches YYYY-MM-DD
    pct = as.numeric(pct),
    biden_approval = as.numeric(`Do you approve or disapprove of the job Joe Biden is doing as President? Approve`)
  ) %>%
  filter(end_date >= as.Date("2024-01-01"))  # Limit data to 2024 and onward

# Separate data for Democrats, Republicans, and Biden Approval
democrat_data <- time_data_graph %>%
  filter(party == "DEM", !is.na(pct)) %>%
  filter(state == "National", !is.na(pct))

republican_data <- time_data_graph %>%
  filter(party == "REP", !is.na(pct)) %>%
  filter(state == "National", !is.na(pct))

biden_approval <- time_data_graph  %>%
  filter(state == "National", !is.na(pct))%>%
  select(end_date, biden_approval) %>%
  filter(!is.na(biden_approval)) %>%
  distinct()

@@

  # Plot Biden's Approval over time (light blue)
  geom_line(data = biden_approval, aes(x = end_date, y = biden_approval, color = "Biden Approval"), size = 1) +

  # Add a vertical line for a key date
  geom_vline(xintercept = as.Date("2024-07-21"), color = "red", linetype = "dashed") +

  # Customize labels, theme, colors, and y-axis limits
  labs(
    title = "Democrat vs. Republican and Biden Approval Over Time",
    x = "Date",
    y = "Percentage (%)",
    color = "Legend"
  ) +
  scale_color_manual(values = c("Democrats" = "blue", "Republicans" = "red", "Biden Approval" = "lightblue")) +
  ylim(25, 55) + # Set y-axis range from 25 to 55, adjust if needed
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

# Model {#sec-model}
## Model Development
The goal of our model is to forecast the popular vote outcome of the 2024 US presidential election, focusing specifically on eight swing states. We will choose the predictor variables based on the correlation matrix we created previously (@fig-predictors-party-pct). It is important to note that even if the percentage of people who vote for Trump (representing the Republican Party) or Harris (representing the Democratic Party) is slightly below 50%, this does not necessarily mean that the corresponding candidate will not be elected as the next President of the United States, given the percentage of people who choose not to vote.

```{r}
#| label: fig-hist-for-republican
#| fig-cap: Distribution of Republican Party Supporting Rate
#| echo: false

# Calculate the variable that represents the Republican Party Supporting Rate
# grouped based on various states
data_rep1 <- poll_data %>%
  group_by(state) %>%
  summarise(
    rep_total_pct = round(sum(.data[[column_refs$republican]], na.rm = TRUE), 1)
  )

# Create a histogram based on targeted data variable
ggplot(data = data_rep1, aes(x=rep_total_pct)) +
  geom_histogram(binwidth = 2) +
  labs(title = "Republic Party Supporting Rate Distirbution", x = "Percentage", y = "Count")

The following @fig-hist-for-democratic is the histogram of the Democratic Party Supporting Rate, which describes the distribution of the response variable for the second model. Our final choice is to construct two generalized linear models: one for the percentage that Trump will become the next President and one for Harris. We can then compare the predicted percentages of people voting for each candidate to make a final prediction of who is more likely to be the next President of the United States.

```{r}
#| label: fig-hist-for-democratic
#| fig-cap: Distribution of Democratic Party Supporting Rate
#| echo: false

# Calculate the variable that represents the Republican Party Supporting Rate
# grouped based on various states
data_rep1 <- poll_data %>%
  group_by(state) %>%
  summarise(
    dem_total_pct = round(sum(.data[[column_refs$democrat]], na.rm = TRUE), 1)
  )

# Create a histogram based on targeted data variable
ggplot(data = data_rep1, aes(x=dem_total_pct)) +
  geom_histogram(binwidth = 2) +
  labs(title = "Democratic Party Supporting Rate Distirbution", x = "Percentage", y = "Count")

To decide between using a generalized linear model or a linear model for each of the two models, we checked the normality of the response variables: the percentage of people supporting the Republican Party and the percentage supporting the Democratic Party. According to @fig-hist-for-democratic and @fig-hist-for-republican, the distributions of both response variables do not appear normal. The distribution of the Democratic Party Supporting Rate tends to resemble a Poisson distribution, with most percentages having a similar frequency and an outlier around 67-68 percent. The Republican Party Supporting Rate distribution has a maximum frequency around 50 percent, with similar frequencies around 25-26 percent and 48-49 percent, suggesting a potential violation of the normal distribution assumption.

Therefore, since the response variables are not normally distributed, we will use a generalized linear model for both models to better capture the characteristics of the data.

The goal of our model is to forecast the likely vote outcome in 8 swing states for the 2024 US presidential election. This model will allow us to estimate the expected vote share for each candidate based on state-level characteristics and prior voting patterns, which are especially important in determining close elections.

## Model Set-up

We define the model as a Bayesian linear model where the outcome, \( y_i \), represents the predicted vote share in state \( i \). The predictors in the model capture key demographic and historical voting patterns across swing states, and we aim to model this outcome with an intercept, \( \alpha \), and a state-specific predictor coefficient, \( \beta_i \), that represents the effect of a given variable \( x_i \) on vote share.

### Model Components and Assumptions

>>>>

# Specify the predictor columns
predictor_columns <- c(
  "What is the highest level of education you have attained? College graduate",
  "What is the highest level of education you have attained? High school or less",
  "Can you please tell me your gender? Female",
  "Can you please tell me your gender? Male",
  "For statistical purposes only, can you please tell me your ethnicity? White or Caucasian",
  "For statistical purposes only, can you please tell me your ethnicity? Hispanic or Latino of any race",
  "Who did you vote for in the 2020 election? Donald Trump",
  "Who did you vote for in the 2020 election? Joe Biden",
  "Although you are undecided, which candidate do you lean toward? Donald Trump",
  "Although you are undecided, which candidate do you lean toward? Kamala Harris"
)

@@

From this analysis, we identified eight critical swing states: Georgia, Michigan, National, Nevada, New Hampshire, North Carolina, Pennsylvania, and Wisconsin. These states will be central to our election prediction. Although Minnesota emerged as a potential swing state, we excluded it due to the wide range of polling data, as explained in @tbl-final-summary. Additionally, we will consider the national average in our overall prediction.

```{r}
#| label: fig-party-results
#| fig-cap: Example Predictor Variable Visualization
#| echo: false
#| warning: false
#| message: false

@@

# Calculate 95% confidence interval for each unique state in 'DEM' data
dem_ci_summary <- dem_data %>%
  group_by(state) %>%
  summarise(
    mean_pct = mean(pct, na.rm = TRUE),
    sd_pct = sd(pct, na.rm = TRUE),
    n = sum(!is.na(pct)),
    lower_95 = ifelse(n >= 2, mean_pct - qt(0.975, df = n - 1) * (sd_pct / sqrt(n)), NA),
    upper_95 = ifelse(n >= 2, mean_pct + qt(0.975, df = n - 1) * (sd_pct / sqrt(n)), NA)
  ) %>%
  # Filter out intervals where the bounds go below 40 or above 60, and remove rows where mean_pct exceeds 50
  filter(!(lower_95 < 35 | upper_95 > 65) & mean_pct <= 100)

@@

# Calculate 95% confidence interval for each unique state in 'REP' data
rep_ci_summary <- rep_data %>%
  group_by(state) %>%
  summarise(
    mean_pct = mean(pct, na.rm = TRUE),
    sd_pct = sd(pct, na.rm = TRUE),
    n = sum(!is.na(pct)),
    lower_95 = ifelse(n >= 2, mean_pct - qt(0.975, df = n - 1) * (sd_pct / sqrt(n)), NA),
    upper_95 = ifelse(n >= 2, mean_pct + qt(0.975, df = n - 1) * (sd_pct / sqrt(n)), NA)
  ) %>%
  # Filter out intervals where the bounds go below 40 or above 60, and remove rows where mean_pct exceeds 50
  filter(!(lower_95 < 35 | upper_95 > 65) & mean_pct <= 50)

@@

```{r}
#| label: fig-party-extremes
#| fig-cap: Demographic Differences Between Maryland and South Dakota
#| echo: false
#| warning: false
#| message: false

# Prepare the poll_summary data frame
poll_summary <- poll_data %>%
  mutate(party_diff = democrat_pct - republican_pct) %>%
  select(-c(2, 3, 4, 6, 7, 8, 9, 11)) %>%  # Remove specific columns by index
  select(-contains('Mode')) %>%             # Remove columns containing 'Mode' in their names
  select(-contains('registration'))

# Function to compute mean differences for all numeric variables
compute_mean_differences <- function(data, blue_state, red_state) {
  numeric_vars <- data %>%
    select(-state, -party_diff) %>%
    select(where(is.numeric)) %>%
    names()

  blue_means <- data %>%
    filter(state == blue_state) %>%
    summarise(across(all_of(numeric_vars), mean, na.rm = TRUE))

  red_means <- data %>%
    filter(state == red_state) %>%
    summarise(across(all_of(numeric_vars), mean, na.rm = TRUE))

@@

  top_diff_vars %>%
    pivot_longer(
      cols = c("Blue_Mean", "Red_Mean"), 
      names_to = "Group", 
      values_to = "Mean"
    ) %>%
    mutate(Group = recode(Group, "Blue_Mean" = "Maryland", "Red_Mean" = "South Dakota"))
}

# Function to create the comparison plot
create_comparison_plot <- function(plot_data) {
  ggplot(plot_data, aes(
    x = reorder(Variable, -Mean), 
    y = Mean, 
    fill = Group, 
    label = round(Mean, 1)
  )) +
    geom_col(position = position_dodge(width = 0.9)) +
    geom_text(
      position = position_dodge(width = 0.9), 
      vjust = -0.3, 
      size = 3
    ) +
    # Update the colors and labels for the groups
    scale_fill_manual(values = c("Maryland" = "blue", "South Dakota" = "red")) +
    labs(
      title = "Demographic Differences Between Maryland and South Dakota",
      x = "Demographic Variable",
      y = "Mean Percentage",
      fill = "State"
    ) +
    scale_y_continuous(limits = c(0, 100)) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 4)
    )
}

# Main analysis function
analyze_state_differences <- function(data, blue_state, red_state, top_n_vars = 4) {
  diff_df <- compute_mean_differences(data, blue_state, red_state)

@@

# Run the analysis for Maryland and South Dakota
analysis_results <- analyze_state_differences(poll_summary, blue_state = "Maryland", red_state = "South Dakota", top_n_vars = 5)

@@

column_refs <- list(
  democrat = "democrat_pct",
  republican = "republican_pct",
  highly_educated = "What is the highest level of education you have attained? College graduate",
  lower_educated = "What is the highest level of education you have attained? High school or less",
  female = "Can you please tell me your gender? Female",
  male = "Can you please tell me your gender? Male",
  caucasian = "For statistical purposes only, can you please tell me your ethnicity? White or Caucasian",
  hispanic_latino = "For statistical purposes only, can you please tell me your ethnicity? Hispanic or Latino of any race",
  vote_trump = "Who did you vote for in the 2020 election? Donald Trump",
  vote_biden = "Who did you vote for in the 2020 election? Joe Biden",
  lean_trump ="Although you are undecided, which candidate do you lean toward? Donald Trump",
  lean_Kamala ="Although you are undecided, which candidate do you lean toward? Kamala Harris"

@@

# Calculate poll_summary with rounding to the nearest tenth
poll_summary <- poll_data %>%
  group_by(state) %>%
  summarise(
    dem_total_pct = round(sum(.data[[column_refs$democrat]], na.rm = TRUE), 1),
    rep_total_pct = round(sum(.data[[column_refs$republican]], na.rm = TRUE), 1),
    highly_educated_pct = round(mean(.data[[column_refs$highly_educated]], na.rm = TRUE), 1),
    lower_educated_pct = round(mean(.data[[column_refs$lower_educated]], na.rm = TRUE), 1),
    female_pct = round(mean(.data[[column_refs$female]], na.rm = TRUE), 1),
    male_pct = round(mean(.data[[column_refs$male]], na.rm = TRUE), 1),
    lean_trump_pct = round(mean(.data[[column_refs$lean_trump]], na.rm = TRUE), 1),
    lean_kamala_pct = round(mean(.data[[column_refs$lean_Kamala]], na.rm = TRUE), 1),
    caucasian_pct = round(mean(.data[[column_refs$caucasian]], na.rm = TRUE), 1),
    minority_pct = round(100 - caucasian_pct, 1)
  ) %>%
  mutate(party_diff = round(dem_total_pct - rep_total_pct, 1)) %>%
  drop_na()

# Select relevant numeric columns for correlation matrix, excluding PartyDifference
corr_data <- poll_summary %>%
  select(
    Democrat = dem_total_pct,
    Republican = rep_total_pct,
    HighlyEducated = highly_educated_pct,
    LowerEducated = lower_educated_pct,
    Female = female_pct,
    Male = male_pct,
    LeanTrump = lean_trump_pct,
    LeanKamala = lean_kamala_pct,
    Caucasian = caucasian_pct,
    Minority = minority_pct
  )

@@

# Visualize the correlation matrix without hierarchical clustering
ggcorrplot(cor_matrix, 
           hc.order = FALSE,  
           type = "lower", 
           lab = TRUE, 
           lab_size = 2,  # Size for correlation labels
           title = "Correlation Matrix of Predictors and Party Lean",
           colors = c("blue", "white", "red"),  
           legend.title = "Correlation") +
  theme(
    plot.title = element_text(size = 12, hjust = 0.5),  # Title size and centered
    axis.text = element_text(size = 1),  # Axis text size
    legend.text = element_text(size = 10),  # Legend text size
    legend.title = element_text(size = 10),  # Legend title size
    axis.title = element_text(size = 1)  # Axis labels size
  )

```{r}
#| label: fig-predictors-party-pct-2
#| fig-cap: Example Voting and Leaning Predictor Variable Visualization
#| echo: false
#| warning: false
#| message: false

# Calculate poll_summary_2 with rounding to the nearest tenth
poll_summary_2 <- poll_data %>%
  group_by(state) %>%
  summarise(
    dem_total_pct = round(sum(.data[[column_refs$democrat]], na.rm = TRUE), 1),
    rep_total_pct = round(sum(.data[[column_refs$republican]], na.rm = TRUE), 1),
    vote_trump_pct = round(mean(.data[[column_refs$vote_trump]], na.rm = TRUE), 1),
    vote_biden_pct = round(mean(.data[[column_refs$vote_biden]], na.rm = TRUE), 1),
    lean_trump_pct = round(mean(.data[[column_refs$lean_trump]], na.rm = TRUE), 1),
    lean_kamala_pct = round(mean(.data[[column_refs$lean_Kamala]], na.rm = TRUE), 1)
  ) %>%
  mutate(party_diff = round(dem_total_pct - rep_total_pct, 1)) %>%
  drop_na()

# Select relevant numeric columns for correlation matrix, excluding PartyDifference
corr_data_2 <- poll_summary_2 %>%
  select(
    Democrat = dem_total_pct,
    Republican = rep_total_pct,
    VotedTrump2020 = vote_trump_pct,
    VotedBiden2020 = vote_biden_pct,
    LeanTrump = lean_trump_pct,
    LeanKamala = lean_kamala_pct
  )

@@

The correlation matrix in @fig-predictors-party-pct shows how different predictors relate to party preferences. We can see which factors, like education or gender, are more strongly tied to Democratic or Republican leanings. Blue shading indicates stronger Democratic support, while red shows stronger Republican support. Key observations include higher education being linked to Democratic support and lower education favoring Trump in 2020. Minority groups generally lean Democratic, while Caucasian/White voters tend to support Trump. Past voter behavior shows Biden supporters remain Democratic, and Trump supporters disapprove of Biden. Additionally, women lean slightly more Democratic compared to men, who often lean Republican.

```{r}
#| label: fig-points-over-time
#| tbl-cap: "Parties Approval and Leaning Preferences"
#| echo: false
#| warning: false
#| message: false

# Convert `end_date` to Date format, ensure numeric columns, and limit to 2024 and onward
time_data_graph <- time_data %>%
  mutate(
    end_date = as.Date(end_date, format = "%Y-%m-%d"),  # Format matches YYYY-MM-DD
    pct = as.numeric(pct),
    lean_trump = as.numeric(`Although you are undecided, which candidate do you lean toward? Donald Trump`),
    lean_kamala = as.numeric(`Although you are undecided, which candidate do you lean toward? Kamala Harris`)
  ) %>%
  filter(end_date >= as.Date("2024-01-01")) %>%  # Limit data to 2024 and onward
  arrange(end_date)  # Ensure data is sorted by end_date

# Separate data for Democrats and Republicans
democrat_data <- time_data_graph %>%
  filter(party == "DEM", !is.na(pct)) %>%
  filter(state == "National") %>%
  arrange(end_date)

republican_data <- time_data_graph %>%
  filter(party == "REP", !is.na(pct)) %>%
  filter(state == "National") %>%
  arrange(end_date)

# Average `lean_trump` values for the same date, keep one entry, and limit to 4 entries per month
lean_trump_data <- time_data_graph %>%
  select(end_date, lean_trump) %>%
  filter(!is.na(lean_trump)) %>%
  group_by(month = floor_date(end_date, "month")) %>%
  arrange(end_date) %>%
  slice_head(n = 4) %>%  # Limit to 4 entries per month
  ungroup() %>%
  select(-month) %>%
  group_by(end_date) %>%
  summarise(lean_trump = mean(lean_trump, na.rm = TRUE)) %>%  # Average values
  arrange(end_date)

# Average `lean_kamala` values for the same date, keep one entry, and limit to 4 entries per month
lean_kamala_data <- time_data_graph %>%
  select(end_date, lean_kamala) %>%
  filter(!is.na(lean_kamala)) %>%
  group_by(month = floor_date(end_date, "month")) %>%
  arrange(end_date) %>%
  slice_head(n = 4) %>%  # Limit to 4 entries per month
  ungroup() %>%
  select(-month) %>%
  group_by(end_date) %>%
  summarise(lean_kamala = mean(lean_kamala, na.rm = TRUE)) %>%  # Average values
  arrange(end_date)

@@

  # Plot Leaning Toward Trump over time (dark orange)
  geom_line(data = lean_trump_data, aes(x = end_date, y = lean_trump, color = "Lean Trump"), size = 0.6) +

  # Plot Leaning Toward Kamala over time (purple)
  geom_line(data = lean_kamala_data, aes(x = end_date, y = lean_kamala, color = "Lean Kamala"), size = 0.6) +

  # Customize labels, theme, colors, and y-axis limits
  labs(
    title = "Democrat vs. Republican and Leaning Preferences Over Time",
    x = "Date",
    y = "Percentage (%)",
    color = "Legend"
  ) +
  scale_color_manual(values = c("Democrats" = "blue", "Republicans" = "red", "Lean Trump" = "darkorange", "Lean Kamala" = "purple")) +
  ylim(25, 55) +  # Set y-axis range from 25 to 55, adjust if needed
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

# Model {#sec-model}
## Model Development

### Forecasting Vote Outcomes in Key Swing States

Our model aims to forecast the voting outcome of the 2024 U.S. presidential election, focusing specifically on eight pivotal swing states. We carefully selected predictor variables based on a correlation matrix we generated previously (see Figure @fig-predictors-party-pct).

### Model Design and Approach

We constructed two separate Bayesian generalized linear models to estimate the percentage likelihood of Trump and Harris winning in each swing state. Our models look at demographic and historical data that let us to generate probabilistic predictions for both candidates. These models provide insights into which candidate is more likely to secure a win, even in closely contested states.

### Key Components of Our Models

- **Outcome Variables**: The percentage of the popular vote each candidate is expected to receive.
- **Predictor Variables**: We include factors such as education level (e.g., college graduate vs. high school or less), gender distribution (male vs. female), and ethnic demographics (White or Caucasian vs. Hispanic or Latino). Additionally, historical voting patterns (e.g., 2020 voting preferences) and current leanings toward Trump or Harris are crucial predictors.
- **Model Structure**: Each model incorporates an intercept to represent baseline support and predictor coefficients to quantify the effect of each variable on the vote share. We use a Beta distribution to account for the bounded nature of percentage outcomes, ensuring our predictions remain between 0 and 100%.

### Assumptions Underlying Our Models

1. **Linearity**: The relationship between each predictor and vote share is assumed to be approximately linear.
2. **Independence**: We assume that vote share predictions for each state are independent, conditional on the included predictors.
3. **Non-Normality**: Given the non-normal distribution of our response variables, using a generalized linear model is more appropriate. Our analysis revealed that the distribution of Democratic support resembles a Poisson-like distribution, while Republican support shows multimodal characteristics.
4. **Priors**: We applied weakly informative priors to regularize our predictions, with specific priors for those who voted for Trump and Biden in 2020, aligning with historical voting trends.

### Bayesian Modeling Implementation

We implemented our models using the `brms` package in R, which allows for flexible Bayesian regression modeling. The models were fit with four chains, running 6,000 iterations each, to ensure convergence and reliable posterior estimates. Sensible default priors were adjusted to reflect our prior knowledge about voting behavior, and we conducted posterior predictive checks to assess the models' performance.

### Results and Interpretation

The results, summarized in Table @tbl-modelresults, highlight the predicted vote shares for each candidate in the swing states. For each state, we report the posterior mean, as well as 95% credible intervals, to capture the range of plausible outcomes. Our findings underscore the impact of demographic and historical factors on voting behavior, providing a nuanced understanding of which states are likely to be battlegrounds in the upcoming election.

By using a Bayesian approach, we not only make point estimates but also quantify the uncertainty surrounding these predictions. This is crucial for election forecasting, where even small shifts in voter behavior can have significant implications. Our analysis will inform campaign strategies and identify states where mobilizing specific demographics could be pivotal.

```{r}
#| label: fig-results-model
#| tbl-cap: "Forecasting model results based on poll data"
#| echo: false
#| warning: false
#| message: false
# Define the swing states of interest
swing_states <- c("Michigan", "Georgia", "Nevada", "North Carolina", "New Hampshire", "Wisconsin", "Pennsylvania", "Arizona", "National")
